---
title: "GO and Pathway Analysis"
author: "Núria Paisano Cabrera"
date: "March-April 2024"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

### Software used

- R v4.3.3

Packages:

- gprofiler2 0.2.3
- ggplot2 3.5.0
- dplyr 1.1.4

Based on: <http://127.0.0.1:18001/library/gprofiler2/doc/gprofiler2.html>


## LOAD PACKAGES

If we are installing packages for the first time:
```{r}
if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install("DESeq2")
```

Loading packages:
```{r}
lapply(c("gprofiler2", "dplyr"), library, character.only=TRUE)
  #gprofiler2: GO terms enrichment analysis
  #dplyr: allows data manipulation
```


## FILE PREPARATION

Creation of an overall id file filtered by genes that have at least 1 count:
```{r}
ids <- read.csv2("/media/sbeato/NURIA/TFG/AA_RNAseq/bioinformatic_analysis/AA_DEF/pathway_analysis/hg_CombinedAnnotation_seabassV1.0_ids.csv", sep = ",")
ids_all <- subset(ids, X.1 != "-")   #we eliminate the NA values for clearer input

RNAcounts_ids <- RNAcounts[which(rowSums(RNAcounts)>=1), ]
RNAcounts_idsnames <- rownames(RNAcounts_ids)
ids_all <- subset(ids_all, X %in% RNAcounts_idsnames)
ids_all <- ids_all$X.1
```


After DEA we need to separate the over-expressed and under-expressed genes of each comparison and save them in two separate files:
```{r}
#MvF:
MvF_over <- subset(MvF_res, MvF_res$log2FoldChange > 0)
MvFover <- rownames(MvF_over)
ids_MvF_over <- subset(ids, X %in% MvFover)
write.csv(as.data.frame(ids_MvF_over),file="/media/sbeato/NURIA/TFG/AA_RNAseq/bioinformatic_analysis/AA_DEF/pathway_analysis/ids_MvF_over.csv")
ids_MvF_over_cut <- subset(ids_MvF_over, X.1 != "-")
ids_MvF_over_cut <- ids_MvF_over_cut$X.1
cat(ids_MvF_over_cut,file="/media/sbeato/NURIA/TFG/AA_RNAseq/bioinformatic_analysis/AA_DEF/pathway_analysis/MvF_ids_over.txt",sep="\n")

MvF_under <- subset(MvF_res, MvF_res$log2FoldChange < 0)
MvFunder <- rownames(MvF_under)
ids_MvF_under <- subset(ids, X %in% MvFunder)
write.csv(as.data.frame(ids_MvF_under),file="/media/sbeato/NURIA/TFG/AA_RNAseq/bioinformatic_analysis/AA_DEF/pathway_analysis/ids_MvF_under.csv")
ids_MvF_under_cut <- subset(ids_MvF_under, X.1 != "-")
ids_MvF_under_cut <- ids_MvF_under_cut$X.1
cat(ids_MvF_under_cut,file="/media/sbeato/NURIA/TFG/AA_RNAseq/bioinformatic_analysis/AA_DEF/pathway_analysis/MvF_ids_under.txt",sep="\n")


#NMvF:
NMvF_over <- subset(NMvF_res, NMvF_res$log2FoldChange > 0)
NMvFover <- rownames(NMvF_over)
ids_NMvF_over <- subset(ids, X %in% NMvFover)
write.csv(as.data.frame(ids_NMvF_over),file="/media/sbeato/NURIA/TFG/AA_RNAseq/bioinformatic_analysis/AA_DEF/pathway_analysis/ids_NMvF_over.csv")
ids_NMvF_over_cut <- subset(ids_NMvF_over, X.1 != "-")
ids_NMvF_over_cut <- ids_NMvF_over_cut$X.1
cat(ids_NMvF_over_cut,file="/media/sbeato/NURIA/TFG/AA_RNAseq/bioinformatic_analysis/AA_DEF/pathway_analysis/NMvF_ids_over.txt",sep="\n")

NMvF_under <- subset(NMvF_res, NMvF_res$log2FoldChange < 0)
NMvFunder <- rownames(NMvF_under)
ids_NMvF_under <- subset(ids, X %in% NMvFunder)
write.csv(as.data.frame(ids_NMvF_under),file="/media/sbeato/NURIA/TFG/AA_RNAseq/bioinformatic_analysis/AA_DEF/pathway_analysis/ids_NMvF_under.csv")
ids_NMvF_under_cut <- subset(ids_NMvF_under, X.1 != "-")
ids_NMvF_under_cut <- ids_NMvF_under_cut$X.1
cat(ids_NMvF_under_cut,file="/media/sbeato/NURIA/TFG/AA_RNAseq/bioinformatic_analysis/AA_DEF/pathway_analysis/NMvF_ids_under.txt",sep="\n")

```

There is no transformation needed for the NMvM comparison since there are only 3 DEGs, and thus a GO analysis will be impossible.


## GO ANALYSIS - GENE LIST FUNCTIONAL ENRICHMENT

### CONSIDERATIONS

Regarding the R instructions:
- The query input is a vector type `c()` containing the GO IDs, or directly the name of an already saved vector. Set `multi_query = TRUE` to input multiple vectors.
- Only set `ordered_query = TRUE` if we know of an already existing hierarchical relation between genes, otherwise set it to `FALSE`.
- If `significant = FALSE`, we must also set `highlight = FALSE`.
- By default, the overrepresented genes are measured. To visualize the underrepresented genes, set `measure_underrepresentation = TRUE`.
- The parameter `user_threshold` corresponds to the p-value used. In this case it is set to 0.01.
- The developers of the package recommend using `correction_method = "g_SCS"`, but given that the rest of our analysis has used FDR, we will use that as well for this part (`correction_method = "fdr"`).
- By default, the `domain_scope` is set to show all annotated genes of the specified organism. To input a custom list, we must set `domain_scope = "custom_annotated"` and add in `custom_bg` the name of a vector with the gene list we want to use as background (in our case, background corresponds to the GO terms of all genes that got at least 1 read in the RNAseq).

In this analysis, we will use as reference organisms _Homo sapiens_ (hsapiens) and _Danio rerio_ (drerio) first since they are model organisms which are better annotated.

Due to the very few DEGs obtained in the NMvM comparison, we can only do a GO terms analysis on the MvF and NMvF comparisons.

Results should also be converted into dataframes and saved for future reference.


### DANIO RERIO

```{r}
#MvF overexpressed:
gostres_MvF_up_dr <- gost(query = ids_MvF_over_cut, 
                            organism = "drerio", ordered_query = FALSE, 
                            multi_query = FALSE, significant = TRUE, exclude_iea = FALSE, 
                            measure_underrepresentation = FALSE, evcodes = FALSE,
                            user_threshold = 0.01, correction_method = "fdr", 
                            domain_scope = "custom_annotated", custom_bg = ids_all, 
                            numeric_ns = "", sources = NULL, as_short_link = FALSE, highlight = TRUE)

gostres_dataframe <- gostres_MvF_up_dr$result %>%
  as.data.frame
gostres_dataframe$parents <- as.character(gostres_dataframe$parents)
gostres_dataframe <- gostres_dataframe[order(gostres_dataframe$p_value),]
write.csv(as.data.frame(gostres_dataframe), file="/media/sbeato/NURIA/TFG/AA_RNAseq/bioinformatic_analysis/AA_DEF/pathway_analysis/gProfiler/def_dataframes/MvF_drerio_up.csv")

gostplot(gostres_MvF_up_dr, capped = TRUE, interactive = TRUE)

p <- gostplot(gostres_MvF_up_dr, capped = TRUE, interactive = FALSE)
publish_gostplot(p, highlight_terms = c("GO:0016020", "GO:0032502", "GO:0048856"), 
                       width = NA, height = NA, filename = "/media/sbeato/NURIA/TFG/AA_RNAseq/bioinformatic_analysis/AA_DEF/pathway_analysis/gProfiler/def_graphs/MvF_drerio_up.png")
  #Add only the GO term numbers we wish to check
  #To check the GO term number, use the dataframe result of the gostres object and search for name of term in question
```

```{r}
#MvF underexpressed:
gostres_MVF_down_dr <- gost(query = ids_MvF_under_cut, 
                             organism = "drerio", ordered_query = FALSE, 
                             multi_query = FALSE, significant = TRUE, exclude_iea = FALSE, 
                             measure_underrepresentation = FALSE, evcodes = FALSE,
                             user_threshold = 0.01, correction_method = "fdr", 
                             domain_scope = "custom_annotated", custom_bg = ids_all, 
                             numeric_ns = "", sources = NULL, as_short_link = FALSE, highlight = TRUE)

gostres_dataframe <- gostres_MVF_down_dr$result %>%
  as.data.frame
gostres_dataframe$parents <- as.character(gostres_dataframe$parents)
gostres_dataframe <- gostres_dataframe[order(gostres_dataframe$p_value),]
write.csv(as.data.frame(gostres_dataframe), file="/media/sbeato/NURIA/TFG/AA_RNAseq/bioinformatic_analysis/AA_DEF/pathway_analysis/gProfiler/def_dataframes/MvF_drerio_down.csv")

gostplot(gostres_MVF_down_dr, capped = TRUE, interactive = TRUE)

p <- gostplot(gostres_MVF_down_dr, capped = TRUE, interactive = FALSE)
publish_gostplot(p, highlight_terms = c("GO:0038023", "GO:0060089", "GO:0004888", "GO:0023052", "GO:0007154"), 
                       width = NA, height = NA, filename = "/media/sbeato/NURIA/TFG/AA_RNAseq/bioinformatic_analysis/AA_DEF/pathway_analysis/gProfiler/def_graphs/MvF_drerio_down.png")
```

```{r}
#NMvF overexpressed:
gostres_NMvF_up_dr <- gost(query = ids_NMvF_over_cut, 
                            organism = "drerio", ordered_query = FALSE, 
                            multi_query = FALSE, significant = TRUE, exclude_iea = FALSE, 
                            measure_underrepresentation = FALSE, evcodes = FALSE,
                            user_threshold = 0.01, correction_method = "fdr", 
                            domain_scope = "custom_annotated", custom_bg = ids_all, 
                            numeric_ns = "", sources = NULL, as_short_link = FALSE, highlight = TRUE)

gostres_dataframe <- gostres_NMvF_up_dr$result %>%
  as.data.frame
gostres_dataframe$parents <- as.character(gostres_dataframe$parents)
gostres_dataframe <- gostres_dataframe[order(gostres_dataframe$p_value),]
write.csv(as.data.frame(gostres_dataframe), file="/media/sbeato/NURIA/TFG/AA_RNAseq/bioinformatic_analysis/AA_DEF/pathway_analysis/gProfiler/def_dataframes/NMvF_drerio_up.csv")

gostplot(gostres_NMvF_up_dr, capped = TRUE, interactive = TRUE)

p <- gostplot(gostres_NMvF_up_dr, capped = TRUE, interactive = FALSE)
publish_gostplot(p, highlight_terms = c("GO:0016020", "GO:0032502", "GO:0048856"), 
                       width = NA, height = NA, filename = "/media/sbeato/NURIA/TFG/AA_RNAseq/bioinformatic_analysis/AA_DEF/pathway_analysis/gProfiler/def_graphs/NMvF_drerio_up.png")
```

```{r}
#NMvF underexpressed:
gostres_NMvF_down_dr <- gost(query = ids_NMvF_under_cut, 
                             organism = "drerio", ordered_query = FALSE, 
                             multi_query = FALSE, significant = TRUE, exclude_iea = FALSE, 
                             measure_underrepresentation = FALSE, evcodes = FALSE,
                             user_threshold = 0.001, correction_method = "fdr", 
                             domain_scope = "custom_annotated", custom_bg = ids_all, 
                             numeric_ns = "", sources = NULL, as_short_link = FALSE, highlight = TRUE)

gostres_dataframe <- gostres_NMvF_down_dr$result %>%
  as.data.frame
gostres_dataframe$parents <- as.character(gostres_dataframe$parents)
gostres_dataframe <- gostres_dataframe[order(gostres_dataframe$p_value),]
write.csv(as.data.frame(gostres_dataframe), file="/media/sbeato/NURIA/TFG/AA_RNAseq/bioinformatic_analysis/AA_DEF/pathway_analysis/gProfiler/def_dataframes/NMvF_drerio_down.csv")

gostplot(gostres_NMvF_down_dr, capped = TRUE, interactive = TRUE)

p <- gostplot(gostres_NMvF_down_dr, capped = TRUE, interactive = FALSE)
publish_gostplot(p, highlight_terms = c("GO:0038023", "GO:0060089", "GO:0004888", "GO:0023052", "GO:0007154"), 
                       width = NA, height = NA, filename = "/media/sbeato/NURIA/TFG/AA_RNAseq/bioinformatic_analysis/AA_DEF/pathway_analysis/gProfiler/def_graphs/NMvF_drerio_down.png")
```

```{r}
#Multi-query overexpressed:
multi_gostres_up_dr <- gost(query = list("MvF" = c(ids_MvF_over_cut),
                             "NMvF" = c(ids_NMvF_over_cut)), 
                             organism = "drerio", ordered_query = FALSE, 
                             multi_query = TRUE, significant = TRUE, exclude_iea = FALSE, 
                             measure_underrepresentation = FALSE, evcodes = FALSE, 
                             user_threshold = 0.01, correction_method = "fdr", 
                             domain_scope = "custom_annotated", custom_bg = ids_all, 
                             numeric_ns = "", sources = c("GO:MF", "GO:CC", "GO:BP", "KEGG", "REAC", "WP"), as_short_link = FALSE, highlight = TRUE)

multi_gostres_dataframe <- multi_gostres_up_dr$result %>%
  as.data.frame
multi_gostres_dataframe$p_values <- as.character(multi_gostres_dataframe$p_values)
multi_gostres_dataframe$significant <- as.character(multi_gostres_dataframe$significant)
multi_gostres_dataframe$query_sizes <- as.character(multi_gostres_dataframe$query_sizes)
multi_gostres_dataframe$intersection_sizes <- as.character(multi_gostres_dataframe$intersection_sizes)
multi_gostres_dataframe$parents <- as.character(multi_gostres_dataframe$parents)
multi_gostres_dataframe$highlighted <- as.character(multi_gostres_dataframe$highlighted)
write.csv(as.data.frame(multi_gostres_dataframe), file="/media/sbeato/NURIA/TFG/AA_RNAseq/bioinformatic_analysis/AA_DEF/pathway_analysis/gProfiler/def_dataframes/multi_drerio_up.csv")

gostplot(multi_gostres_up_dr, capped = TRUE, interactive = TRUE)

p <- gostplot(multi_gostres_up_dr, capped = TRUE, interactive = FALSE)
publish_gostplot(p, highlight_terms = c("GO:0016020", "GO:0032502", "GO:0048856"), 
                       width = NA, height = NA, filename = "/media/sbeato/NURIA/TFG/AA_RNAseq/bioinformatic_analysis/AA_DEF/pathway_analysis/gProfiler/def_graphs/multi_drerio_up.png")
```

```{r}
#Multi-query underexpressed:
multi_gostres_down_dr <- gost(query = list("MvF" = c(ids_MvF_under_cut),
                             "NMvF" = c(ids_NMvF_under_cut)), 
                             organism = "drerio", ordered_query = FALSE, 
                             multi_query = TRUE, significant = TRUE, exclude_iea = FALSE, 
                             measure_underrepresentation = FALSE, evcodes = FALSE, 
                             user_threshold = 0.01, correction_method = "fdr", 
                             domain_scope = "custom_annotated", custom_bg = ids_all,
                             numeric_ns = "", sources = c("GO:CC", "GO:BP", "KEGG", "REAC", "TF", "WP"), as_short_link = FALSE, highlight = TRUE)

multi_gostres_dataframe <- multi_gostres_down_dr$result %>%
  as.data.frame
multi_gostres_dataframe$p_values <- as.character(multi_gostres_dataframe$p_values)
multi_gostres_dataframe$significant <- as.character(multi_gostres_dataframe$significant)
multi_gostres_dataframe$query_sizes <- as.character(multi_gostres_dataframe$query_sizes)
multi_gostres_dataframe$intersection_sizes <- as.character(multi_gostres_dataframe$intersection_sizes)
multi_gostres_dataframe$parents <- as.character(multi_gostres_dataframe$parents)
multi_gostres_dataframe$highlighted <- as.character(multi_gostres_dataframe$highlighted)
write.csv(as.data.frame(multi_gostres_dataframe), file="/media/sbeato/NURIA/TFG/AA_RNAseq/bioinformatic_analysis/AA_DEF/pathway_analysis/gProfiler/def_dataframes/multi_drerio_down.csv")

gostplot(multi_gostres_down_dr, capped = TRUE, interactive = TRUE)

p <- gostplot(multi_gostres_down_dr, capped = TRUE, interactive = FALSE)
publish_gostplot(p, highlight_terms = c("GO:0038023", "GO:0060089", "GO:0004888", "GO:0023052", "GO:0007186"), 
                       width = NA, height = NA, filename = "/media/sbeato/NURIA/TFG/AA_RNAseq/bioinformatic_analysis/AA_DEF/pathway_analysis/gProfiler/def_graphs/multi_drerio_down.png")

```


### HOMO SAPIENS

```{r}
#MvF overexpressed:
gostres_MvF_up_hs <- gost(query = ids_MvF_over_cut, 
                            organism = "hsapiens", ordered_query = FALSE, 
                            multi_query = FALSE, significant = TRUE, exclude_iea = FALSE, 
                            measure_underrepresentation = FALSE, evcodes = FALSE,
                            user_threshold = 0.01, correction_method = "fdr", 
                            domain_scope = "custom_annotated", custom_bg = ids_all, 
                            numeric_ns = "", sources = NULL, as_short_link = FALSE, highlight = TRUE)

gostres_dataframe <- gostres_MvF_up_hs$result %>%
  as.data.frame
gostres_dataframe$parents <- as.character(gostres_dataframe$parents)
  #the column "parents" needs to be set to character (instead of list) in order to write the dataframe
gostres_dataframe <- gostres_dataframe[order(gostres_dataframe$p_value),]
write.csv(as.data.frame(gostres_dataframe), file="/media/sbeato/NURIA/TFG/AA_RNAseq/bioinformatic_analysis/AA_DEF/pathway_analysis/gProfiler/def_dataframes/MvF_hsapiens_up.csv")

gostplot(gostres_MvF_up_hs, capped = TRUE, interactive = TRUE)
  #NOTE: if saved as png, the graph is no longer interactive

p <- gostplot(gostres_MvF_up_hs, capped = TRUE, interactive = FALSE)
publish_gostplot(p, highlight_terms = c("GO:0016020", "GO:0032502", "GO:0048856", "GO:0110165", "GO:0005488"), 
                       width = NA, height = NA, filename = "/media/sbeato/NURIA/TFG/AA_RNAseq/bioinformatic_analysis/AA_DEF/pathway_analysis/gProfiler/def_graphs/MvF_hsapiens_up.png")
  #Add only the GO term numbers we wish to check
  #To check the GO term number, use the dataframe result of the gostres object and search for name of term in question
```

```{r}
#MvF underexpressed:
gostres_MvF_down_hs <- gost(query = ids_MvF_under_cut, 
                             organism = "hsapiens", ordered_query = FALSE, 
                             multi_query = FALSE, significant = TRUE, exclude_iea = FALSE, 
                             measure_underrepresentation = FALSE, evcodes = FALSE,
                             user_threshold = 0.01, correction_method = "fdr", 
                             domain_scope = "custom_annotated", custom_bg = ids_all, 
                             numeric_ns = "", sources = NULL, as_short_link = FALSE, highlight = TRUE)

gostres_dataframe <- gostres_MVF_down_hs$result %>%
  as.data.frame
gostres_dataframe$parents <- as.character(gostres_dataframe$parents)
gostres_dataframe <- gostres_dataframe[order(gostres_dataframe$p_value),]
write.csv(as.data.frame(gostres_dataframe), file="/media/sbeato/NURIA/TFG/AA_RNAseq/bioinformatic_analysis/AA_DEF/pathway_analysis/gProfiler/def_dataframes/MvF_hsapiens_down.csv")

gostplot(gostres_MVF_down_hs, capped = TRUE, interactive = TRUE)

p <- gostplot(gostres_MVF_under_hs, capped = TRUE, interactive = FALSE)
publish_gostplot(p, highlight_terms = c("GO:0045291", "GO:0000353", "GO:0000365"), 
                       width = NA, height = NA, filename = "/media/sbeato/NURIA/TFG/AA_RNAseq/bioinformatic_analysis/AA_DEF/pathway_analysis/gProfiler/def_graphs/MvF_hsapiens_down.png")
```

```{r}
#NMvF overexpressed:
gostres_NMvF_up_hs <- gost(query = ids_NMvF_over_cut, 
                            organism = "hsapiens", ordered_query = FALSE, 
                            multi_query = FALSE, significant = TRUE, exclude_iea = FALSE, 
                            measure_underrepresentation = FALSE, evcodes = FALSE,
                            user_threshold = 0.01, correction_method = "fdr", 
                            domain_scope = "custom_annotated", custom_bg = ids_all, 
                            numeric_ns = "", sources = NULL, as_short_link = FALSE, highlight = TRUE)

gostres_dataframe <- gostres_NMvF_up_hs$result %>%
  as.data.frame
gostres_dataframe$parents <- as.character(gostres_dataframe$parents)
gostres_dataframe <- gostres_dataframe[order(gostres_dataframe$p_value),]
write.csv(as.data.frame(gostres_dataframe), file="/media/sbeato/NURIA/TFG/AA_RNAseq/bioinformatic_analysis/AA_DEF/pathway_analysis/gProfiler/def_dataframes/NMvF_hsapiens_up.csv")

gostplot(gostres_NMvF_up_hs, capped = TRUE, interactive = TRUE)

p <- gostplot(gostres_NMvF_up_hs, capped = TRUE, interactive = FALSE)
publish_gostplot(p, highlight_terms = c("GO:0016020", "GO:0032502", "GO:0048856", "GO:0110165", "GO:0005488"), 
                       width = NA, height = NA, filename = "/media/sbeato/NURIA/TFG/AA_RNAseq/bioinformatic_analysis/AA_DEF/pathway_analysis/gProfiler/def_graphs/NMvF_hsapiens_up.png")
```

```{r}
#NMvF underexpressed:
gostres_NMvF_down_hs <- gost(query = ids_NMvF_under_cut, 
                             organism = "hsapiens", ordered_query = FALSE, 
                             multi_query = FALSE, significant = TRUE, exclude_iea = FALSE, 
                             measure_underrepresentation = FALSE, evcodes = FALSE,
                             user_threshold = 0.01, correction_method = "fdr", 
                             domain_scope = "custom_annotated", custom_bg = ids_all, 
                             numeric_ns = "", sources = NULL, as_short_link = FALSE, highlight = TRUE)

gostres_dataframe <- gostres_NMvF_down_hs$result %>%
  as.data.frame
gostres_dataframe$parents <- as.character(gostres_dataframe$parents)
gostres_dataframe <- gostres_dataframe[order(gostres_dataframe$p_value),]
write.csv(as.data.frame(gostres_dataframe), file="/media/sbeato/NURIA/TFG/AA_RNAseq/bioinformatic_analysis/AA_DEF/pathway_analysis/gProfiler/def_dataframes/NMvF_hsapiens_down.csv")

gostplot(gostres_NMvF_down_hs, capped = TRUE, interactive = TRUE)

p <- gostplot(gostres_NMvF_down_hs, capped = TRUE, interactive = FALSE)
publish_gostplot(p, highlight_terms = c("GO:0045291", "GO:0000353", "GO:0000365"), 
                       width = NA, height = NA, filename = "/media/sbeato/NURIA/TFG/AA_RNAseq/bioinformatic_analysis/AA_DEF/pathway_analysis/gProfiler/def_graphs/NMvF_hsapiens_down.png")
```

```{r}
#Multi-query overexpressed:
multi_gostres_up_hs <- gost(query = list("MvF" = c(ids_MvF_over_cut),
                             "NMvF" = c(ids_NMvF_over_cut)), 
                             organism = "hsapiens", ordered_query = FALSE, 
                             multi_query = TRUE, significant = TRUE, exclude_iea = FALSE, 
                             measure_underrepresentation = FALSE, evcodes = FALSE, 
                             user_threshold = 0.01, correction_method = "fdr", 
                             domain_scope = "custom_annotated", custom_bg = ids_all, 
                             numeric_ns = "", sources = c("GO:MF", "GO:CC", "GO:BP", "KEGG", "REAC", "TF", "MIRNA", "HPA", "HP", "WP"), as_short_link = FALSE, highlight = TRUE)

multi_gostres_dataframe <- multi_gostres_up_hs$result %>%
  as.data.frame
multi_gostres_dataframe$p_values <- as.character(multi_gostres_dataframe$p_values)
multi_gostres_dataframe$significant <- as.character(multi_gostres_dataframe$significant)
multi_gostres_dataframe$query_sizes <- as.character(multi_gostres_dataframe$query_sizes)
multi_gostres_dataframe$intersection_sizes <- as.character(multi_gostres_dataframe$intersection_sizes)
multi_gostres_dataframe$parents <- as.character(multi_gostres_dataframe$parents)
multi_gostres_dataframe$highlighted <- as.character(multi_gostres_dataframe$highlighted)
write.csv(as.data.frame(multi_gostres_dataframe), file="/media/sbeato/NURIA/TFG/AA_RNAseq/bioinformatic_analysis/AA_DEF/pathway_analysis/gProfiler/def_dataframes/multi_hsapiens_up.csv")

gostplot(multi_gostres_up_hs, capped = TRUE, interactive = TRUE)

p <- gostplot(multi_gostres_up_hs, capped = TRUE, interactive = FALSE)
publish_gostplot(p, highlight_terms = c("GO:0016020", "GO:0032502", "GO:0048856", "GO:0110165", "GO:0005488"), 
                       width = NA, height = NA, filename = "/media/sbeato/NURIA/TFG/AA_RNAseq/bioinformatic_analysis/AA_DEF/pathway_analysis/gProfiler/def_graphs/multi_hsapiens_up.png")
```

```{r}
#Multi-query underexpressed:
multi_gostres_down_hs <- gost(query = list("MvF" = c(ids_MvF_under_cut),
                             "NMvF" = c(ids_NMvF_under_cut)), 
                             organism = "hsapiens", ordered_query = FALSE, 
                             multi_query = TRUE, significant = TRUE, exclude_iea = FALSE, 
                             measure_underrepresentation = FALSE, evcodes = FALSE, 
                             user_threshold = 0.01, correction_method = "fdr", 
                             domain_scope = "custom_annotated", custom_bg = ids_all, 
                             numeric_ns = "", sources = NULL, as_short_link = FALSE, highlight = TRUE)

multi_gostres_dataframe <- multi_gostres_down_hs$result %>%
  as.data.frame
multi_gostres_dataframe$p_values <- as.character(multi_gostres_dataframe$p_values)
multi_gostres_dataframe$significant <- as.character(multi_gostres_dataframe$significant)
multi_gostres_dataframe$query_sizes <- as.character(multi_gostres_dataframe$query_sizes)
multi_gostres_dataframe$intersection_sizes <- as.character(multi_gostres_dataframe$intersection_sizes)
multi_gostres_dataframe$parents <- as.character(multi_gostres_dataframe$parents)
multi_gostres_dataframe$highlighted <- as.character(multi_gostres_dataframe$highlighted)
write.csv(as.data.frame(multi_gostres_dataframe), file="/media/sbeato/NURIA/TFG/AA_RNAseq/bioinformatic_analysis/AA_DEF/pathway_analysis/gProfiler/def_dataframes/multi_hsapiens_down.csv")

gostplot(multi_gostres_down_hs, capped = TRUE, interactive = TRUE)

p <- gostplot(multi_gostres_down_hs, capped = TRUE, interactive = FALSE)
publish_gostplot(p, highlight_terms = c("GO:0045291", "GO:0000353", "GO:0000365"), 
                       width = NA, height = NA, filename = "/media/sbeato/NURIA/TFG/AA_RNAseq/bioinformatic_analysis/AA_DEF/pathway_analysis/gProfiler/def_graphs/multi_hsapiens_down.png")

```


### ANALYSIS OF KEGG GENES

List of KEGG genes of the two different upregulated pathways:
```{r}
gostres_MvF_up_hs <- gost(query = ids_MvF_over_cut, 
                            organism = "hsapiens", ordered_query = FALSE, 
                            multi_query = FALSE, significant = TRUE, exclude_iea = FALSE, 
                            measure_underrepresentation = FALSE, evcodes = TRUE,
                            user_threshold = 0.01, correction_method = "fdr", 
                            domain_scope = "custom_annotated", custom_bg = ids_all, 
                            numeric_ns = "", sources = "KEGG", as_short_link = FALSE, highlight = TRUE)

gostres_dataframe <- gostres_MvF_up_hs$result %>%
  as.data.frame
gostres_dataframe$parents <- as.character(gostres_dataframe$parents)
gostres_dataframe <- gostres_dataframe[order(gostres_dataframe$p_value),]
gostres_dataframe <- select(gostres_dataframe,-evidence_codes)
write.csv(as.data.frame(gostres_dataframe), file="/media/sbeato/NURIA/TFG/AA_RNAseq/bioinformatic_analysis/AA_DEF/pathway_analysis/gProfiler/def_dataframes/MvF_hsapiens_up_KEGGgenes.csv")
```

```{r}
gostres_NMvF_up_hs <- gost(query = ids_NMvF_over_cut, 
                            organism = "hsapiens", ordered_query = FALSE, 
                            multi_query = FALSE, significant = TRUE, exclude_iea = FALSE, 
                            measure_underrepresentation = FALSE, evcodes = TRUE,
                            user_threshold = 0.01, correction_method = "fdr", 
                            domain_scope = "custom_annotated", custom_bg = ids_all, 
                            numeric_ns = "", sources = "KEGG", as_short_link = FALSE, highlight = TRUE)

gostres_dataframe <- gostres_NMvF_up_hs$result %>%
  as.data.frame
gostres_dataframe$parents <- as.character(gostres_dataframe$parents)
gostres_dataframe <- gostres_dataframe[order(gostres_dataframe$p_value),]
gostres_dataframe <- select(gostres_dataframe,-evidence_codes)
write.csv(as.data.frame(gostres_dataframe), file="/media/sbeato/NURIA/TFG/AA_RNAseq/bioinformatic_analysis/AA_DEF/pathway_analysis/gProfiler/def_dataframes/NMvF_hsapiens_up_KEGGgenes.csv")
```

Extracting the genes belonging to the pathways found in NMvF and not MvF (gap junction, cytokine-cytokine receptor interaction) and comparing their presence in other lists:
```{r}
gap_junction <- gostres_dataframe[26,15]   #ensure that we are using the dataframe with the NMvF comparison results!
cytokine <- gostres_dataframe[27,15]

#Creation of the gene vectors:
gap_junction <- strsplit(gap_junction, ",") %>% unlist(gap_junction)
cytokine <- strsplit(cytokine, ",") %>% unlist(cytokine)

#COMPARISONS:
#MvF:
found_up_gj <- intersect(gap_junction,ids_MvF_over_cut)
found_up_c <- intersect(cytokine,ids_MvF_over_cut)

notfound_up_gj <- setdiff(gap_junction,ids_MvF_over_cut)
notfound_up_c <- setdiff(cytokine,ids_MvF_over_cut)

found_down_gj <- intersect(gap_junction,ids_MvF_under_cut)
found_down_c <- intersect(cytokine,ids_MvF_under_cut)

notfound_down_gj <- setdiff(gap_junction,ids_MvF_under_cut)
notfound_down_c <- setdiff(cytokine,ids_MvF_under_cut)

#NMvF:
found_up_gj <- intersect(gap_junction,ids_NMvF_over_cut)   #vector should have 30 values
found_up_c <- intersect(cytokine,ids_NMvF_over_cut)   #vector should have 49 values

notfound_up_gj <- setdiff(gap_junction,ids_NMvF_over_cut)
notfound_up_c <- setdiff(cytokine,ids_NMvF_over_cut)   #both of these vectors should be empty

found_down_gj <- intersect(gap_junction,ids_NMvF_under_cut)
found_down_c <- intersect(cytokine,ids_NMvF_under_cut)

notfound_down_gj <- setdiff(gap_junction,ids_NMvF_under_cut)
notfound_down_c <- setdiff(cytokine,ids_NMvF_under_cut)
```
