---
title: "Graphical analysis"
author: "Núria Paisano Cabrera"
date: "March 2024"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

### Software used

- R v4.3.3

Packages:

- DESeq2 1.40.2
- dplyr 1.1.4
- ggplot2 3.5.0
- ggrepel 0.9.5
- tidyverse 2.0.0
- vsn 3.68.0
- EnhancedVolcano 1.18.0


## LOAD PACKAGES

If we are installing packages for the first time:
```{r}
if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install("packagename")
```

Loading packages:
```{r}
lapply(c("DESeq2", "dplyr", "ggplot2", "ggrepel", "tidyverse", "vsn", "EnhancedVolcano"), library, character.only=TRUE)
  #DESeq2: differential expression analysis
  #vsn: extraction of normalized vaules
  #dplyr/tidyverse: allows data manipulation
  #ggplot2/ggrepel/EnhancedVolcano: graphs
```

_NOTE: Sometimes the ggplot package gives an error if DESeq2 is loaded. If this happens unload DESeq2:_
```{r}
detach("package:DESeq2", unload=TRUE)
```


## MA-PLOT

Transform results into dataframe:
```{r}
SeabassFM_DE <- results(SeabassFM_DE, contrast = c("Sex",'M','F'), alpha=0.01) %>%
  as.data.frame

SeabassNMM_DE <- results(SeabassNMM_DE, contrast = c("Sex",'NM','M'), alpha=0.01) %>%
  as.data.frame

SeabassFNM_DE <- results(SeabassFNM_DE, contrast = c("Sex",'NM','F'), alpha=0.01) %>%
  as.data.frame
```

MA plots:
```{r}
iv.sig <- SeabassFM_DE$padj < 0.01
iv.up <- SeabassFM_DE$log2FoldChange > 2 & iv.sig
iv.dn <- SeabassFM_DE$log2FoldChange < -2 & iv.sig
plot(log2(SeabassFM_DE$baseMean + 1), SeabassFM_DE$log2FoldChange, pch=20, col="grey",
     main="Males vs Females", xlab="log2(baseMean)", ylab="Log2FC") + points(log2(SeabassFM_DE$baseMean + 1)[iv.up], SeabassFM_DE$log2FoldChange[iv.up], col="blue", pch=20) + points(log2(SeabassFM_DE$baseMean + 1)[iv.dn], SeabassFM_DE$log2FoldChange[iv.dn], col="red3", pch=20) + abline(h=c(-2,2), col="cyan4", lwd=2) + abline(h=0, col="gray50", lwd=3)

iv.sig <- SeabassNMM_DE$padj < 0.01
iv.up <- SeabassNMM_DE$log2FoldChange > 2 & iv.sig
iv.dn <- SeabassNMM_DE$log2FoldChange < -2 & iv.sig
plot(log2(SeabassNMM_DE$baseMean + 1), SeabassNMM_DE$log2FoldChange, pch=20, col="grey",
     main="Neomales vs Males", xlab="log2(baseMean)", ylab="Log2FC") + points(log2(SeabassNMM_DE$baseMean + 1)[iv.up], SeabassNMM_DE$log2FoldChange[iv.up], col="blue", pch=20) + points(log2(SeabassNMM_DE$baseMean + 1)[iv.dn], SeabassNMM_DE$log2FoldChange[iv.dn], col="red3", pch=20) + abline(h=c(-2,2), col="cyan4", lwd=2) + abline(h=0, col="gray50", lwd=3)

iv.sig <- SeabassFNM_DE$padj < 0.01
iv.up <- SeabassFNM_DE$log2FoldChange > 2 & iv.sig
iv.dn <- SeabassFNM_DE$log2FoldChange < -2 & iv.sig
plot(log2(SeabassFNM_DE$baseMean + 1), SeabassFNM_DE$log2FoldChange, pch=20, col="grey",
     main="Neomales vs Females", xlab="log2(baseMean)", ylab="Log2FC") + points(log2(SeabassFNM_DE$baseMean + 1)[iv.up], SeabassFNM_DE$log2FoldChange[iv.up], col="blue", pch=20) + points(log2(SeabassFNM_DE$baseMean + 1)[iv.dn], SeabassFNM_DE$log2FoldChange[iv.dn], col="red3", pch=20) + abline(h=c(-2,2), col="cyan4", lwd=2) + abline(h=0, col="gray50", lwd=3)
```

**REMEMBER TO SAVE THE GRAPHS AS IMAGE FILES (EG JPEG)**


## MODEL DISPERSION PLOT

Extracting transformed (normalized) values:
```{r}
vsd <- vst(SeabassFNM_DE, blind=FALSE)
rld <- rlog(SeabassFNM_DE, blind=FALSE)
head(assay(vsd), 3)
```

Effects of transformation on the variance (this gives log2(n+1)):
```{r}
ntd <- normTransform(SeabassFNM_DE)

meanSdPlot(assay(ntd), xlab = "mean ntd", ylab = "sd ntd")

meanSdPlot(assay(vsd), xlab = "mean vsd", ylab = "sd vsd")

meanSdPlot(assay(rld), xlab = "mean rld", ylab = "sd rld")
```

Model dispersion plot:
```{r}
plotDispEsts(SeabassFM_DE, main = "Males vs Females")
plotDispEsts(SeabassNMM_DE, main = "Neomales vs Males")
plotDispEsts(SeabassFNM_DE, main = "Neomales vs Females")
```

**REMEMBER TO SAVE THE GRAPHS AS IMAGE FILES (EG JPEG)**


## VOLCANO PLOT

```{r}
library(EnhancedVolcano)

EnhancedVolcano(SeabassFM_DE,
                lab = rownames(SeabassFM_DE),
                x = 'log2FoldChange',
                y = 'padj',
                axisLabSize = 11.0,
                title = 'Male vs Female',
                subtitleLabSize = 0.0,
                caption = bquote(~Log[2]~ "fold change cutoff, 2; p-value cutoff, 0,01"),
                captionLabSize = 11.0,
                pCutoff = 0.01,   #pvalue of 0.01
                FCcutoff = 2,   #Fold Change of 2
                pointSize = 1.0,
                labSize = 0.0,
                legendLabels=c('Not sig.', 'Log2 FC', 'p-value', 'p-value & Log2 FC'),
                legendPosition = 'right',
                legendLabSize = 8.0,
                legendIconSize = 3.0)

EnhancedVolcano(SeabassNMM_DE,
                lab = rownames(SeabassNMM_DE),
                x = 'log2FoldChange',
                y = 'padj',
                axisLabSize = 11.0,
                title = 'Neomale vs Male',
                subtitleLabSize = 0.0,
                caption = bquote(~Log[2]~ "fold change cutoff, 2; p-value cutoff, 0,01"),
                captionLabSize = 11.0,
                pCutoff = 0.01,
                FCcutoff = 2,
                pointSize = 1.0,
                labSize = 0.0,
                legendLabels=c('Not sig.', 'Log2 FC', 'p-value', 'p-value & Log2 FC'),
                legendPosition = 'right',
                legendLabSize = 8.0,
                legendIconSize = 3.0)

EnhancedVolcano(SeabassFNM_DE,
                lab = rownames(SeabassFNM_DE),
                x = 'log2FoldChange',
                y = 'padj',
                axisLabSize = 11.0,
                title = 'Neomale vs Female',
                subtitleLabSize = 0.0,
                caption = bquote(~Log[2]~ "fold change cutoff, 2; p-value cutoff, 0,01"),
                captionLabSize = 8.0,
                pCutoff = 0.01,
                FCcutoff = 2,
                pointSize = 1.0,
                labSize = 0.0,
                legendLabels=c('Not sig.', 'Log2 FC', 'p-value', 'p-value & Log2 FC'),
                legendPosition = 'right',
                legendLabSize = 8.0,
                legendIconSize = 3.0)
```

**REMEMBER TO SAVE THE GRAPHS AS IMAGE FILES (EG JPEG)**


## HIGHEST LOG FOLD CHANGE VISUALIZATION
```{r}
ggplot(MvF_overexpr, aes(log2(padj), log2FoldChange)) +
  geom_point(aes(colour = MvF_overexpr$`ids_MvF$X.1`), size = 2.5) +
  theme(legend.position="none") + 
  expand_limits(x=c(100,-600), y=c(5,15)) +
  geom_label_repel(
    label=MvF_overexpr$`ids_MvF$X.1`,
    aes(log2(padj), log2FoldChange, colour = MvF_overexpr$`ids_MvF$X.1`),
    max.overlaps=Inf,
    force = 30,
    segment.color = "grey80"
  )

ggplot(MvF_underexpr, aes(log2(padj), log2FoldChange)) +
  geom_point(aes(colour = MvF_underexpr$`ids_MvF$X.1`), size = 2.5) +
  theme(legend.position="none") + 
  expand_limits(x=c(0,-600), y=c(-12, -9)) +
  geom_label_repel(
    label=MvF_underexpr$`ids_MvF$X.1`,
    aes(log2(padj), log2FoldChange, colour = MvF_underexpr$`ids_MvF$X.1`),
    max.overlaps=Inf,
    force = 40,
    segment.color = "grey80"
  ) +
  scale_y_continuous(trans = "reverse")

ggplot(NMvF_overexpr, aes(log2(padj), log2FoldChange)) +
  geom_point(aes(colour = NMvF_overexpr$`ids_NMvF$X.1`), size = 2.5) +
  theme(legend.position="none") + 
  expand_limits(x=c(100,-600), y=c(5,15)) +
  geom_label_repel(
    label=NMvF_overexpr$`ids_NMvF$X.1`,
    aes(log2(padj), log2FoldChange, colour = NMvF_overexpr$`ids_NMvF$X.1`),
    max.overlaps=Inf,
    force = 30,
    segment.color = "grey80"
  )

ggplot(NMvF_underexpr[-c(12),], aes(log2(padj), log2FoldChange)) +
  geom_point(aes(colour = NMvF_underexpr[-c(12),]$`ids_NMvF$X.1`), size = 2.5) +
  theme(legend.position="none") + 
  expand_limits(x=c(0,-800), y=c(-12, -8)) +
  geom_label_repel(
    label=NMvF_underexpr[-c(12),]$`ids_NMvF$X.1`,
    aes(log2(padj), log2FoldChange, colour = NMvF_underexpr[-c(12),]$`ids_NMvF$X.1`),
    max.overlaps=Inf,
    force = 40,
    segment.color = "grey80"
  ) +
  scale_y_continuous(trans = "reverse")

ggplot(NMvM_overexpr, aes(log2(padj), log2FoldChange)) +
  geom_point(aes(colour = NMvM_overexpr$`ids_NMvM$X.1`), size = 2.5) +
  theme(legend.position="none") + 
  geom_label_repel(
    label=NMvM_overexpr$`ids_NMvM$X.1`,
    aes(log2(padj), log2FoldChange, colour = NMvM_overexpr$`ids_NMvM$X.1`),
    max.overlaps=Inf,
    force = 50,
    segment.color = "grey80"
  )

```

**REMEMBER TO SAVE THE GRAPHS AS IMAGE FILES (EG JPEG)**

## GO ENRICHMENT SCATTERPLOT

MvF homo sapiens up:
```{r}
KEGGM <- gostres_MvF_up_hs$result %>%
  as.data.frame
KEGGM <- filter(KEGGM, KEGGM$source == "KEGG")

ggplot(KEGGM, aes(x=recall, y=term_name, color= KEGGM$p_value, size = KEGGM$intersection_size)) + 
  geom_point() + 
  labs(colour = "p-value",
       size = "Number of genes") + 
  theme(legend.position = "right") +
  ggtitle("Upregulated genes in Males vs Females") + 
  xlab("Enrichment factor") + ylab("KEGG pathway")
```

NMvF homo sapiens up:
```{r}
KEGGNM <- gostres_NMvF_up_hs$result %>%
  as.data.frame
KEGGNM <- filter(KEGGNM, KEGGNM$source == "KEGG")

ggplot(KEGGNM, aes(x=recall, y=term_name, color= KEGGNM$p_value, size = KEGGNM$intersection_size)) + 
  geom_point() + 
  labs(colour = "p-value",
       size = "Number of genes") + 
  theme(legend.position = "right") +
  ggtitle("Upregulated genes in Neomales vs Females") + 
  xlab("Enrichment factor") + ylab("KEGG pathway")
```
